Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-06-03 SIMPaper/SIMPaper.Rmd
finalized version for arxiv

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
revised: 2022-06-01
accepted: 2022-07-01
abstract: "In survival analysis, covariate interactions are difficult to model using regression-based methods without prior knowledge. Neural network-based survival models have been developed for data-driven interaction modeling, however, these approaches lack implementations for time-varying interactions or a flexible baseline hazard. To address this, we extend the case-base sampling framework and propose Case-Base Neural Networks (CBNN), which introduce time directly into the model, using commonly available neural network components for a simple implementation. First, we case-base sample the survival data, converting the survival time of each participant into discrete moments in time. This allows us to estimate the probability an event occurs at a given moment using a neural network to get the hazard. We compare CBNN to survival methods based on regression and neural networks in two simulations and two real data applications with a fixed neural network architecture. We report metrics for each model over follow-up time using both a proper (Index of Prediction Accuracy) and non-proper (inverse probability censoring weights-adjusted concordance index) scoring rule. A simple simulation shows CBNN outperforming the other neural network methods. The complex simulation highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, outperforming all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. CBNN has shown that it is more consistent in prediction performance across time in the simulations and real data applications compared to the competing neural network approaches."

@@

In survival analysis, covariate interactions are difficult to model using regression-based methods without prior knowledge. Neural network-based survival models have been developed for data-driven interaction modeling, however, these approaches can not model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) which extends the case-base sampling framework which introduce time directly into the model. We case-base sample the survival data, converting the survival time of each participant into discrete moments in time. This allows us to use a feed-forward neural network that estimate the probability of an event occurring at a given moment, which can be used to estimate the hazard function. We then compare CBNN to survival methods based on regression and neural networks in two simulations and two real data applications with a fixed neural network architecture. We report metrics for each model over follow-up time using both proper (Index of Prediction Accuracy) and non-proper (inverse probability censoring weights-adjusted concordance index) scoring rules. A simple simulation with an exponential hazard model shows CBNN consistently outperforming the other neural network methods. The complex simulation highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, outperforming all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time than the competing neural network approaches.

# Introduction

what is aft, dont like assuming distribution, smooth. Less used than cox. Proportional hazards assumptions hard to maintain in long study, all regression require prior knowledge.

The widespread use of the Cox proportional hazards model has led to a focus on survival models based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009].
In clinical settings, Cox proportional hazards models are frequently prioritized over smooth-in-time accelerated failure time (AFT) models [@hanley2009], which can estimate absolute risks by modeling the hazard directly through user specification of a baseline hazard distribution. Using splines as the baseline hazard is an alternative to assuming a specific distribution [@royston2002flexible]. The risk associated with each covariate may not be constant over time, making it difficult to maintain the proportional hazards assumption. Previous studies related to breast cancer have discovered time-varying interactions with covariates of interest, like tumor size [@coradini2000time]. A limitation of regression-based models is the requirement of prior knowledge; the analyst must know potential time-varying interactions and explicitly evaluate them.

Smooth-in-time accelerated failure time (AFT) models can estimate absolute risks by modeling the hazard directly through user specification of a baseline hazard distribution [@kleinbaum2012survival]. The assumption of a specific distribution tends to be a point of contention. A method has been developped using splines, providing flexibility to the baseline hazard  [@royston2002flexible]. AFT models are underutilized compared to Cox proportional hazards, resulting in a focus on survival models based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009]. However, The risk associated with each covariate may not be constant over time, making it difficult to maintain the proportional hazards assumption. Previous studies related to breast cancer have discovered time-varying interactions with covariates of interest, like tumor size [@coradini2000time]. A limitation of all regression-based models is the requirement of prior knowledge; the analyst must know potential time-varying interactions and explicitly evaluate them.

Neural networks provide a data-driven approach to approximating interaction terms. DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood as a custom loss function [@katzman2018DeepSurv]. This results in a stepwise absolute risk curve that cannot accommodate time-varying interactions. Compared to Cox regression, DeepSurv has shown an improvement in performance on the Study to Understand Prognoses Preferences and Risk Treatments (SUPPORT) dataset [@knaus1995SUPPORT]. Farraggi et al. suggested a modification to the loss function required for a Cox neural network that can handle non-proportional hazards [@faraggi1995neural]. Wang et al. took the concept of extreme learning machines and applied them in a survival context, removing the need for tuning the number of layers and nodes [@wang2019extreme].

DeepHit requires that each survival time of interest be specified in the model and directly estimates survival curves, rather than deriving a hazard function [@lee2018DeepHit]. It assumes an inverse Gaussian distribution as the baseline hazard and its concordance index (c-index) score outperforms DeepSurv [@katzman2018DeepSurv] on the Molecular Taxonomy of Breast Cancer International Consortium (METABRIC) dataset [@curtis2012genomic].

Deep Survival Machines (DSM) provides a flexible, parametric survival model using neural networks with a mixture of distributions as the baseline hazard [@dsmPaper]. On both the SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit [@dsmPaper]. However, like DeepHit and DeepSurv, DSM cannot model time-varying interactions.

These methods need custom loss and activation functions that are difficult to implement in practice. We note the complexity of these alternative neural network approaches; DeepSurv, DeepHit, and DSM all require custom loss functions [@katzman2018DeepSurv] [@lee2018DeepHit] [@dsmPaper], while DSM requires a two-phase learning process and user implementations for distributions beyond Log-Normal or Weibull [@dsmPaper]. Regression-based approaches require prior specification of all interaction terms, which makes it challenging to model covariate effects that change over time. In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying interactions and a flexible baseline hazard using commonly available neural network components. Our approach to modeling the full hazard uses case-base sampling [@hanley2009]. This sampling technique allows probabilistic models to predict survival outcomes. Following the case-base framework, we can use transformations of time as a feature (covariate) to specify different baseline hazards. For example, by including splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model [@royston2002flexible], however, this still requires user specification of time-varying interactions. CBNN can model both without user specification; only a sufficiently deep network is required.

In Section \ref{methods}, we describe how case-base sampling and neural networks are combined both conceptually and algebraically, along with our hyperparameter choices and software implementation. In Section \ref{sims}, we describe our metrics and compare the performance of CBNN, DeepSurv, DeepHit, DSM, Cox regression, and case-base using logistic regression on simulated data. Section \ref{casestudies} describes the real-data analysis, while in Section \ref{discussion}, we explore the implications of our results and contextualize them within neural network survival analysis in a single event setting.

# Case-base neural network methodology, metrics, and software {#methods}

In this section, we define case-base sampling, which converts the total survival time into discrete person-specific moments (person-moments). Then, we detail how neural networks can be used within this framework, explicitly incorporating time as a feature while adjusting for the sampling bias. Finally, we report on the software versions used. A package is available for use at [GITHUBLINK]. The entire code base to generate the analyses in this paper is available at [GITHUB LINK].

@@

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By allowing a neural network to approximate a higher-order polynomial of time, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of nodes, and layers control the flexibility of the model. We provide a detailed description of the choices we made in the next sub-section.

@@

Neural networks provide flexibility when defining the architecture and optimization parameters of our models. These hyperparameter choices can affect the estimated parameters and were chosen during a set of initial simulations to determine if CBNN can learn complex interactions in practice. We set a max number of epochs to be $2000$, batch size as $512$, learning rate as $10^{-3}$, decay as $10^{-7}$, patience as $10$ epochs, $\{50,50,25,25\}$ nodes in each hidden layer with $50\%$ dropout at each layer, a validation split of $20\%$, a testing split of $10%$, a minimum delta loss on the validation set of $10^{-7}$ over 10 epochs, and Adam [@gulli2017] as our optimizer. These choices may permit the model to approximate higher-order interactions while preventing over-fitting [@srivastava2014dropout]. We fix a train-validation-test split that allows us to update the weights with a subset of the data (training), assess performance at each epoch (validation) and gauge performance of the final model (test). We select the best weights after training based on validation loss [@gulli2017].

@@

R [@Rsoft] and python [@py] are used to evaluate methods from both languages. We fit the Cox model using the **survival** package [@survpkg], the CBLR model using the **casebase** package [@cbpkg], the DeepSurv model using the **survivalmodels** package [@survmods], the DSM model using **DeepSurvivalMachines** [@dsmPaper] and the DeepHit model using **pycox** [@lee2018DeepHit]. We made the components of CBNN using the **keras** [@keras] package and the **casebase** package for the sampling step. The **simsurv** package [@simsurv] is used to simulate our simulation studies, while **flexsurv** [@flexsurv] is used to fit a flexible baseline hazard using splines for our complex simulation. We use the implementation of c-ipcw from the python package **sksurv** [@sksurv]. The **riskRegression** package [@riskRegression] is used to get the Brier score and IPA metrics. We changed the **riskRegression** package to be used with any user supplied risk function $F$. To ensure that both R and Python-based models are running in unison on the same data through our simulations and bootstrap, we use the **reticulate** package [@reticulate].

@@

We simulate data from a simple model that depends on a constant baseline hazard:
$$h(t \mid X_i) = \lambda\cdot e^{\beta_{{1}}z_{1}+\beta_{{2}}z_{2}+\beta_{{3}}z_{3}},$$
with minimal covariate effects given by $\beta_{{1}}=0.1, \beta_{{2}}=0.1, \beta_{{3}}=0.1$ and $\lambda=1.0$. Once we simulate survival times, we introduce 10% random censoring.

@@

Figure \ref{fig:megaPlot} A, B and Table \ref{tab:megaTable} A show the results for the simple simulation. The regression-based methods (CBLR, Cox, Optimal) outperform the neural network ones in the simple simulation setting. Among the neural network approaches, CBNN outperforms all other methods in terms of both IPA and c-ipcw (Figure \ref{fig:megaPlot} A, B). Specifically, we see CBNN is consistent across time with smaller confidence intervals compared to DeepHit, DeepSurv and DSM.

In this simple setting, the regression models are much closer to the Optimal model, while the neural network models perform worse than the null model. The wide confidence bands in c-ipcw suggest the neural network models may be over-parameterized.

@@

This simulation demonstrates performance with the presence of a complex baseline hazard and a time-varying interaction. Originally used to demonstrate the spline-based hazard model proposed by Royston and Parmar [@royston2002flexible], the breast cancer dataset provides a complex hazard from which we simulate, available in the **flexsurv** R package [@flexsurv]. To increase the complexity of our data-generating mechanism for this simulation, we design the model as follows:
\begin{align}
h(t \mid X_i) =\sum_{i=1}^{5} (\gamma_{i} \cdot \psi_{i}) + \beta_{{1}} (z_{1}) + \beta_{{2}} (z_{2})+ \beta_{{3}} (z_{3})+ \tau_{1} ( z_{1} \cdot z_{2} \cdot t)+ \tau_{2} ( z_{1} \cdot z_{3})+ \tau_{3} (z_{2} \cdot z_{3}), \nonumber
\end{align}

```{=latex}
\FloatBarrier
\begin{table}
\caption{Four tables representing performance at certain follow-up times for the simple simulation, complex simulation, SUPPORT and METABRIC. Each table shows performance for each method in each study at $25\%$, $50\%$. $75\%$ and $100\%$ of follow-up time. The bold elements show the best model for each study, at each follow-up time of interest. These tables are included to provide exact measures at certain intervals. The models of interest are: Cox, case-base with logistic regression (CBLR), DeepSurv, DeepHit, Case-Base Neural Networks (CBNN), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Deep Survival Machines (DSM).}
\label{tab:megaTable}
```

```{r megaTable, echo = FALSE , fig.align = 'center', out.width = "100%"}
#readRDS(here::here("figures", "megaPlot.rds"))
knitr::include_graphics(here::here("../pmnnPlayground/reviewedAnalysisPMNN/figures", "megaTable.pdf"))
```

@@

where $\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2$ and $\psi$ are basis splines. To generate the $gamma$ coefficients, let $g$: $\mathbb{R} \rightarrow \mathbb{R}$ be a smoothing method for variable $t$ by a projection on to a set of basis functions on splines: 
\srb{i used gt as the function for splines. Specifically, natural cubic spline function of log time https://www.rdocumentation.org/packages/flexsurv/versions/2.1/topics/flexsurvspline . therefore, should i use g(log(t)) and and replace "smoothing method" with natural cubic spline? SAHIR: will fix any issues in this section in the final version.}
\begin{align}
g(t)=\sum^{m}_{l=1}\psi_{l} (t)\gamma_{l}.
\end{align}

An initial model with three knots ($m=5$) is fit using the *flexsurvspline* function with an intercept-only model on the data, which returns the coefficients ($\gamma$) and basis of time ($\psi$) we use in our model. Note that we fix these values for the analysis. The $\beta$ coefficients represent direct effects, $\tau_{2}$ and ${\tau_3}$ represent interactions and $\tau_{1}$ is a time-varying interaction.

@@

Figure \ref{fig:megaPlot} C, D and Table \ref{tab:megaTable} B show the performance over time on a test set in the complex simulation. Apart from the Optimal regression model, CBNN outperforms the competing models when examining IPA, and up to the 75-percentile of follow-up time for c-ipcw. We expected the Optimal model to perform best in both metrics. However, this was not the case for c-ipcw and may be due to an artifact of concordance-based metrics, where a misspecified model may perform better than a correctly specified one [@cindexfails2019]. We attribute this difference in performance to the flexibility of CBNN in terms of time-varying interactions and baseline hazard, flexibility the other neural network models do not have, explaining their drop in performance.

@@

Our complex simulation demonstrates the superior performance of CBNN in ideal conditions with clean data. To obtain a more realistic performance assessment, we compared models using two real datasets with a time-to-event outcome. The first case study examines the SUPPORT dataset [@knaus1995SUPPORT]. The second case study examines the METABRIC dataset [@curtis2012genomic]. We use the same hyperparameters as in the simulation studies. As we do not know the true model for the real data, we exclude the Optimal model. We split the datasets keeping 20% of the observations as a test set. 20% of the training set is kept aside for validation at each epoch. We predict risk functions for everyone in the test set, which are used to calculate our metrics. We conduct 100 bootstrap re-samples for the real data applications to obtain confidence intervals.

@@

The SUPPORT dataset tracks the time until death for seriously ill patients at five American hospitals [@knaus1995SUPPORT]. We use a pre-processed version of the dataset made available in the DeepSurv package [@katzman2018DeepSurv]. This dataset contains 9104 samples and 14 covariates (age, sex, race, number of comorbidities, presence of diabetes, presence of dementia, presence of cancer, mean arterial blood pressure, heart rate, respiration rate, temperature, white blood cell count, serum’s sodium, and serum’s creatinine) [@katzman2018DeepSurv]. Patients with missing features were excluded and 68.10% of the patients died during the 5.56-year study period [@katzman2018DeepSurv].

Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C demonstrates the performance over time on a test set. The regression models (CBLR, Cox) perform best considering IPA, followed by CBNN from the $25^{th}$ to $100^{th}$ percentile of follow-up time. We note a drop in performance for CBNN before the $25^{th}$ percentile of follow-up. For c-ipcw, CBNN outperforms the competing models consistently over follow-up time. Note that performance is similar for all models, aside from DeepSurv whose c-ipcw is lower than the rest (Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C).

@@

METABRIC is a 30-yearlong study aiming to discover the molecular drivers of breast tumors, following 2000 individuals with breast cancer until death [@curtis2012genomic]. They described these growths as primary invasive breast carcinomas, with the goal of discovering both genetic and clinical risk factors for breast cancer survival [@curtis2012genomic]. We used the processed dataset made available through DeepSurv [@katzman2018DeepSurv], which includes 1980 patients of which 57.72% die due to breast cancer within a median 10 years of follow-up [@katzman2018DeepSurv]. There are 11 covariates in total: 4 RNA-Seq gene expressions (MKI67, EGFR, PGR, and ERBB2) and 5 clinical features (hormone treatment indicator, radiotherapy indicator, chemotherapy indicator, ER-positive indicator, and age at diagnosis) [@katzman2018DeepSurv].

Figure \ref{fig:megaPlot} G, H and Table \ref{tab:megaTable} D shows the performance on a test set over time. The IPA scores suggest that regression models outperform competing models on this dataset, as all the neural network models are equal to or perform worse than KM over follow-up time. Our CBNN model is comparable to KM until around the 50-percentile of follow-up time, after which CBNN, DeepSurv and DSM drop in performance. The c-ipcw produces a different ranking. Our CBNN model outperforms the other models up to the $25^{th}$ percentile of follow-up time, whereas DeepHit performs best from the $50^{th}$ to $75^{th}$ percentile of follow-up. With this metric, CBNN and DeepHit outperform the regression models. This disagreement between IPA and c-ipcw between may be due to the misspecification issue of concordance-based metrics [@cindexfails2019]. The neural network models may be over-parameterized, as shown by the wide confidence bands in c-ipcw.

>>>>

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
revised: 2022-06-01
accepted: 2022-07-01
abstract: "Neural network-based survival methods have been developed to model data-driven covariate interactions. While these methods have led to better predictive performance than regression-based approaches, they cannot model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) as a new approach that combines the case-base sampling framework with flexible architectures. Our method naturally accounts for censoring, does not require additional hyperparameters and uses binary cross-entropy as the loss function. Using a novel sampling scheme and data augmentation, we incorporate time directly into a feed-forward neural network. CBNN predicts the probability of an event occurring at a given moment and estimates the hazard function. We compare the performance of CBNN to survival methods based on regression and neural networks in two simulations and two real data applications. We report two time-dependent metrics for each model. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time and outperforms the competing neural network approaches. For a simple simulation with an exponential hazard model, CBNN outperforms the other neural network methods. For a complex simulation, which highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, CBNN outperforms all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. We highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of survival outcomes. An R package is available at [https://github.com/Jesse-Islam/cbnn](https://github.com/Jesse-Islam/cbnn)."

@@

# Introduction
Smooth-in-time accelerated failure time (AFT) models can estimate absolute risks by modeling the hazard directly through a user-specified baseline hazard distribution [@kleinbaum2012survival]. Cox proportional hazards models are used more often than AFT models, causing analyses to be based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009]. The identification of an appropriate distribution for the baseline hazard in an AFT model may be difficult for common diseases that have many interacting risk factors, or a Cox model where the disease pathogenesis may change with age, making it difficult to maintain the proportional hazards assumption. For example, previous studies of breast cancer incidence have discovered time-varying interactions with covariates of interest, such as tumor size [@coradini2000time]. One approach to provide flexibility in the baseline hazard involves using the basis of splines on time in our model [@royston2002flexible]. However, regression-based models are limited in that they require prior knowledge about potential time-varying interactions and their quantitative effects.

Neural networks provide a data-driven approach to approximating interaction terms. For example, DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood as a custom loss function [@katzman2018DeepSurv], resulting in a stepwise absolute risk curve that cannot accommodate time-varying interactions. Compared to Cox regression, DeepSurv shows better performance on the Study to Understand Prognoses Preferences and Risk Treatments (SUPPORT) dataset [@knaus1995SUPPORT]. To handle non-proportional hazards, a modification to the loss function was proposed [@faraggi1995neural]. To remove the need for tuning the number of layers and nodes, the concept of extreme learning machines has been applied to a Cox neural network model [@wang2019extreme].

As an alternative method that assumes a baseline hazard distribution, DeepHit specifies each survival time of interest in the model and directly estimates survival curves, rather than deriving a hazard function [@lee2018DeepHit]. It assumes an inverse Gaussian distribution as the baseline hazard and it outperforms DeepSurv on the Molecular Taxonomy of Breast Cancer International Consortium (METABRIC) dataset [@curtis2012genomic].

Providing further flexibility, Deep Survival Machines (DSM) is a parametric survival model using neural networks with a mixture of distributions as the baseline hazard [@dsmPaper]. On both the SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit [@dsmPaper]. However, like DeepHit and DeepSurv, DSM cannot model time-varying interactions.

We note that these alternative neural network approaches all require custom loss functions [@katzman2018DeepSurv] [@lee2018DeepHit] [@dsmPaper]. DeepHit introduces a hyperparameter weighing its two loss functions (negative log-likelihood and ranking losses) while DSM requires a two-phase learning process and user implementations for distributions beyond Log-Normal or Weibull  [@lee2018DeepHit] [@dsmPaper]. Regression-based approaches require prior specification of all interaction terms, which makes it challenging to model covariate effects that change over time. The current neural network models provide flexibility at the cost of opacity, while regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying interactions and a flexible baseline hazard using commonly available neural network components. Our approach to modeling the full hazard uses case-base sampling [@hanley2009]. This sampling technique allows probabilistic models to predict survival outcomes. As part of the case-base framework, we use transformations of time as a feature (covariate) to specify different baseline hazards. For example, by including splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model [@royston2002flexible], however, this still requires explicit use of time-varying interactions. CBNN can model both without extra tuning parameters.

In Section \ref{methods}, we describe how case-base sampling and neural networks are combined both conceptually and algebraically, along with our hyperparameter choices and software implementation. Section \ref{sims}, describes our metrics and compares the performance of CBNN, DeepSurv, DeepHit, DSM, Cox regression and case-base using logistic regression (CBLR) on simulated data. Section \ref{casestudies} describes the real-data analysis, while Section \ref{discussion} explores the implications of our results and contextualizes them within neural network survival analysis in a single event setting.

# Case-base neural network methodology, metrics and software {#methods}

In this section, we define case-base sampling, which converts the total survival time into discrete person-specific moments (person-moments). Then, we detail how neural networks can be used within this framework, explicitly incorporating time as a feature while adjusting for the sampling bias. Finally, we report on the software versions used. An R package is available for use at [https://github.com/Jesse-Islam/cbnn](https://github.com/Jesse-Islam/cbnn). The entire code base to reproduce the figures and empirical results in this paper is available at [https://github.com/Jesse-Islam/cbnnManuscript](https://github.com/Jesse-Islam/cbnnManuscript).

@@

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By approximating a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility of the fukk hazard function. We provide a detailed description of the choices we made in the next sub-section.

@@

Neural networks are flexible when defining the architecture and optimization parameters. These hyperparameter decisions can affect the estimated parameters and were chosen during a set of initial simulations to determine if CBNN can learn complex interactions in practice. We set a max number of epochs to be $2000$, batch size as $512$, learning rate as $10^{-3}$, decay as $10^{-7}$, patience as $10$ epochs, $\{50,50,25,25\}$ nodes in each hidden layer with $50\%$ dropout at each layer, a minimum delta loss on the validation set of $10^{-7}$ over 10 epochs and Adam [@gulli2017] as our optimizer. These choices may permit the model to approximate higher-order interactions while preventing over-fitting [@srivastava2014dropout]. We fix a train-validation-test split that allows us to update the weights with a subset of the data (training), assess performance at each epoch (validation) and gauge performance of the final model (test) for each method. We select the best weights after training based on validation loss [@gulli2017].

@@

R [@Rsoft] and python [@py] are used to evaluate methods from both languages. We fit the Cox model using the **survival** package [@survpkg], the CBLR model using the **casebase** package [@cbpkg], the DeepSurv model using the **survivalmodels** package [@survmods], the DeepHit model using **pycox** [@lee2018DeepHit] and the DSM model using **DeepSurvivalMachines** [@dsmPaper]. We made the components of CBNN using the **casebase** package for the sampling step and the **keras** [@keras] package for our neural network architecture. The **simsurv** package [@simsurv] is used for our simulation studies, while **flexsurv** [@flexsurv] is used to fit a flexible baseline hazard using splines for our complex simulation. We use the implementation of $C_{IPCW}$ from the python package **sksurv** [@sksurv]. The **riskRegression** package [@riskRegression] is used to get the Index of Prediction Accuracy (IPA metric). Both metrics are described in detail in the following section. We modify the **riskRegression** package to be used with any user supplied risk function $F$. To ensure that both R and Python-based models are running in unison on the same data through our simulations and bootstrap, we use the **reticulate** package [@reticulate].

@@

We simulate data from a simple model that primarily depends on a constant baseline hazard:
$$\log h(t \mid X_i) = \beta_{{1}}z_{1}+\beta_{{2}}z_{2}+\beta_{{3}}z_{3},$$
The covariate effects are given by $\beta_{{1}}=0.1, \beta_{{2}}=0.1, \beta_{{3}}=0.1$. Once we simulate survival times, we introduce 10% random censoring.

@@

Figure \ref{fig:megaPlot} A, B and Table \ref{tab:megaTable} A show the results for the simple simulation. The regression-based methods (CBLR, Cox, Optimal) outperform the neural network ones in the simple simulation setting. Among the neural network approaches, CBNN outperforms all other methods in terms of both IPA and $C_{IPCW}$ (Figure \ref{fig:megaPlot} A, B). Specifically, we see CBNN is consistent across time with smaller confidence intervals compared to DeepHit, DeepSurv and DSM.

In this simple setting, the regression models are much closer to the Optimal model, while the neural network models perform worse than the KM null model. The wide confidence bands in $C_{IPCW}$ suggest the neural network models may be over-parameterized.

@@

This simulation demonstrates performance with the presence of a complex baseline hazard and a time-varying interaction. Originally used to show the spline-based hazard model proposed by Royston and Parmar [@royston2002flexible], the breast cancer dataset provides a complex hazard from which we simulate, available in the **flexsurv** R package [@flexsurv]. To increase the complexity of our data-generating mechanism for this simulation, we design the model as follows:
\begin{align}
\log h(t \mid X_i) =\sum_{i=1}^{5} (\gamma_{i} \cdot \psi_{i}) + \beta_{{1}} (z_{1}) + \beta_{{2}} (z_{2})+ \beta_{{3}} (z_{3})+ \tau_{1} ( z_{1} \cdot z_{2} \cdot t)+ \tau_{2} ( z_{1} \cdot z_{3})+ \tau_{3} (z_{2} \cdot z_{3}), \nonumber
\end{align}

```{=latex}
\FloatBarrier
\begin{table}
\caption{Four tables representing performance at certain follow-up times for the simple simulation, complex simulation, SUPPORT and METABRIC. Each table shows performance for each method in each study at $25\%$, $50\%$. $75\%$ and $100\%$ of follow-up time. The bold elements show the best model for each study, at each follow-up time of interest. These tables are included to provide exact measures at certain intervals. The models of interest are: Cox, case-base with logistic regression (CBLR), DeepSurv, DeepHit, Case-Base Neural Network (CBNN), Optimal and Deep Survival Machines (DSM).}
\label{tab:megaTable}
```

```{r megaTable, echo = FALSE , fig.align = 'center', out.width = "100%"}
#readRDS(here::here("figures", "megaPlot.rds"))
knitr::include_graphics(here::here("analyses/figures", "megaTable.pdf"))
```

@@

where $\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2$ and $\psi$ are basis splines. The $gamma$ coefficients are obtained from an intercept-only cubic splines model with three knots using the *flexsurvspline* function from the **flexsurv** package [@flexsurv]. Note that we fix these values for the analysis. The $\beta$ coefficients represent direct effects, $\tau_{2}$ and ${\tau_3}$ represent interactions and $\tau_{1}$ is a time-varying interaction.

@@

Figure \ref{fig:megaPlot} C, D and Table \ref{tab:megaTable} B show the performance over time on a test set in the complex simulation. Apart from the Optimal regression model, CBNN outperforms the competing models when examining IPA and up to the 75-percentile of follow-up time for $C_{IPCW}$. We expected the Optimal model to perform best in both metrics. However, this was not the case for $C_{IPCW}$ and may be due to an artifact of concordance-based metrics, where a misspecified model may perform better than a correctly specified one [@cindexfails2019]. We attribute the performance of CBNN to its flexibility in modeling time-varying interactions and baseline hazard, flexibility the other neural network models do not have.

@@

Our complex simulation demonstrates the superior performance of CBNN in ideal conditions with clean data. To obtain a more realistic performance assessment, we compared models using two real datasets with a time-to-event outcome. The first case study examines the SUPPORT dataset [@knaus1995SUPPORT]. The second case study examines the METABRIC dataset [@curtis2012genomic]. We use the same hyperparameters as in the simulation studies. As we do not know the true model for the real data, we exclude the Optimal model. We split the datasets keeping 20% of the observations as a test set. 20% of the training set is kept aside for validation at each epoch. We predict risk functions for everyone in the test set, which is used to calculate our metrics. We conduct 100 bootstrap re-samples for the real data applications to obtain confidence intervals.

@@

The SUPPORT dataset tracks the time until death for seriously ill patients at five American hospitals [@knaus1995SUPPORT]. We use a pre-processed version of the dataset made available in the DeepSurv package [@katzman2018DeepSurv]. This dataset contains 9104 samples and 14 covariates (age, sex, race, number of comorbidities, presence of diabetes, presence of dementia, presence of cancer, mean arterial blood pressure, heart rate, respiration rate, temperature, white blood cell count, serum’s sodium and serum’s creatinine) [@katzman2018DeepSurv]. Patients with missing features were excluded and 68.10% of the patients died during the 5.56-year study period [@katzman2018DeepSurv].

Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C demonstrates the performance over time on a test set. The regression models (CBLR, Cox) perform best considering IPA, followed by CBNN from the $25^{th}$ to $100^{th}$ percentile of follow-up time. We note a drop in performance for CBNN before the $25^{th}$ percentile of follow-up. For $C_{IPCW}$, CBNN outperforms the competing models consistently over follow-up time. Note that performance is similar for all models, aside from DeepSurv whose $C_{IPCW}$ is lower than the rest (Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C).

@@

METABRIC is a 30-yearlong study aiming to discover the molecular drivers of breast tumors, following 2000 individuals with breast cancer until death [@curtis2012genomic]. They described these growths as primary invasive breast carcinomas, with the goal of discovering both genetic and clinical risk factors for breast cancer survival [@curtis2012genomic]. We used the processed dataset made available through DeepSurv [@katzman2018DeepSurv], which includes 1980 patients of which 57.72% die due to breast cancer within a median 10 years of follow-up [@katzman2018DeepSurv]. There are 11 covariates in total: 4 RNA-Seq gene expressions (MKI67, EGFR, PGR and ERBB2) and 5 clinical features (hormone treatment indicator, radiotherapy indicator, chemotherapy indicator, ER-positive indicator and age at diagnosis) [@katzman2018DeepSurv].

Figure \ref{fig:megaPlot} G, H and Table \ref{tab:megaTable} D shows the performance on a test set over time. The IPA scores suggest that regression models outperform competing models on this dataset, as all the neural network models are equal to or perform worse than KM over follow-up time. Our CBNN model is comparable to KM until around the 50-percentile of follow-up time, after which CBNN, DeepSurv and DSM drop in performance. The $C_{IPCW}$ produces a different ranking. Our CBNN model outperforms the other models up to the $25^{th}$ percentile of follow-up time, whereas DeepHit performs best from the $50^{th}$ to $75^{th}$ percentile of follow-up. With this metric, CBNN and DeepHit outperform the regression models. This disagreement between IPA and $C_{IPCW}$ may be due to the misspecification issue of concordance-based metrics [@cindexfails2019]. The neural network models may be over-parameterized, as shown by the wide confidence bands in $C_{IPCW}$.
